---
title: "DESEq2 workflow"
author: Wiktor Mikolajczak
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width  = 5,
  fig.height = 4,
  dpi        = 90,
  fig.align  = "center"
)
```

# Note

Use this vignette as a general guideline, not a ground truth.
There is no one way to analyze genomics data.
Sometimes you have to adjust your analysis based on your research question, experimental design etc.

In the future you will analyze genomic data that you or other people generated.
It is vital that, before any analysis, you familiarize yourself with the data set:

- What is the biological question you are trying to answer?
  What are the hypothesis?

- What is the study design and how was the experiment done (sample type, treatment, genetic knockout, disease, sample size, system, are there any batches etc.)?

- How was the data processed (software, genome version etc.)?

- What type of data are available (raw data, count matrix, sample metadata)?

Instead of just copy-pasting the code, try implementing the following logic:

1.  Read the vignette first. Try to understand each step.
2.  Without looking at the vignette, write down the main steps of the workflow (use headers to build an outline of your document at the same time). For each step, write a short comment: What is this step for?, Why is it important? Compare your answers with the vignette. It is ok if you don't remember the exact code or the name of the functions. It's more important that you understand each step conceptually.
3.  After each step write down your observations. What has changed? Why did you choose this DESeq design? How does PCA explain the data? Are there any batch/technical effects? How many DEGs? etc.
4.  At the end of your analysis, think about and write down conclusions. It doesn't have to be an academic assay but try to state the main points and do a little bit of goggling.
    - Are you able to answer the initial research question and confirm/deny the hypotheses?
    - Try to connect what is known with what you have found.

If you have any questions, check the [DESEq2 vignette](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html).

The goal of this training is to help you understand the general workflow of differential gene expression analysis.
Although using AI can be beneficial and speed up your analysis, it is important that you have a good understanding of topics covered in this training and you have truly mastered the skill.
Only then you can use AI to help you with the analysis.
It is fine, to ask AI to help you with customizing a ggplot, but try not to use it for the actual analysis.

If you don't know what a function does, run `?function_name` in Console.

# About the dataset

This workflow uses the **airway** dataset (Bioconductor package `airway`), an RNA-seq experiment from [Himes et al. 2014, *PLoS ONE*](https://doi.org/10.1371/journal.pone.0099625) (GEO: [GSE52778](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE52778)).
The code used to generate the tables can be found [data/airway_generate_counts.R](https://github.com/wiktormiko/Intro-to-genomics/blob/main/vignettes/data/airway_generate_counts.R)

- **Biological question:** how do human airway smooth muscle (ASM) cells respond to dexamethasone, a glucocorticoid used to treat asthma and inflammation?
- **Design:** four ASM cell lines (`N61311`, `N052611`, `N080611`, `N061011`), each either left untreated (`untrt`) or treated with dexamethasone (`trt`).
- **Size:** 8 samples (4 cell lines x 2 conditions) and 63,677 genes.
- **Count matrix** (`airway_raw_counts.csv`): rows are genes (Ensembl IDs), columns are samples (SRA run IDs, e.g. `SRR1039508`). Values are raw, unnormalized read counts.
- **Metadata** (`airway_metadata.csv`): one row per sample.
  - `cell`: the cell line.
  - `dex`: dexamethasone treatment (`trt` / `untrt`). This is the main variable of interest.
  - `albut`: albuterol treatment. Every sample here is `untrt`, so it carries no information.
  - `SampleName`, `Run`, `Experiment`, `Sample`, `BioSample`: GEO / SRA identifiers.
  - `avgLength`: average read length.

Full description:

```{r, message=FALSE, warning=FALSE}
# library(airway)
# vignette("airway")
```

Because each cell line appears in both conditions, the design is **paired**.
In the DESeq2 model, `cell` should be included as a covariate alongside `dex`.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(annotables)
library(PCAtools)
library(EnhancedVolcano)
library(clusterProfiler)
```

# Import the data

```{r}
raw_counts <- read.csv("data/airway_raw_counts.csv", row.names = 1)
mtd <- read.csv("data/airway_metadata.csv", row.names = 1)
```

```{r}
head(raw_counts)
```

Count matrix is a table where rows correspond to gene names and columns - to samples.
The gene names are encoded in "ENSEMBL" system.
We will convert those names to SYMBOLs - regular gene names.

```{r}
head(mtd)
```

Metadata contains information about samples.
Each row is a sample and each column describes a metric.
For example, we can put information about treatment, batch, collection date, patient ID, etc.
This is a publicly available dataset and the metadata info is shared.
If you perform the experiment yourself, it is important to create such table yourself with as many details about a sample as you can.

It is important that the count table and metadata are matching.
Column names of the count matrix and row names of the metadata have to be the same!

We can ensure those two are matching by running this:

```{r}
all(rownames(mtd) == colnames(raw_counts))
```

if they're not in the right order we can use `match()` to reorder one table so it lines up with the other:

```{r}
raw_counts <- raw_counts[, match(rownames(mtd), colnames(raw_counts))]
```

# Pre–filtering

## Remove low count genes

It is not necessary to filter out low count genes, because DESEq pipeline does it automatically (`independent filtering`), but applying a small cutoff might make the object smaller and require less computational power and memory.
This cutoff is taken from the vignette.
It's also common to use `filterByExp` from edgeR package

`smallestGroupSize` - we have 4 treated and 4 controls so we will use 4.

```{r}
table(mtd$dex)
```

```{r}
ngenes_before <- nrow(raw_counts)
```

```{r}
smallestGroupSize <- 4
keep <- rowSums(raw_counts >= 10) >= smallestGroupSize
raw_counts <- raw_counts[keep,]
```

```{r}
ngenes_after <- nrow(raw_counts)
cat("#genes before filtering: ", ngenes_before, "\n #genes after filtering: ", ngenes_after)
```

1.  `raw_counts >= 10` gives a table of TRUE/FALSE with the same shape as the counts. Each cell asks whether this gene had at least 10 reads in this sample.
2.  `rowSums(...)` adds up the TRUE values for each gene, giving the number of samples where the gene reached 10 reads.
3.  `>= smallestGroupSize` asks whether that number is at least the size of your smallest experimental group.

A gene that is truly different between conditions should have decent counts in all the samples of at least one group.
If your smallest group has 4 samples, requiring at least 4 samples with 10 or more reads means a gene could pass by being expressed in just one condition, such as a gene that is switched on only after treatment.
A gene with high counts in only 1 of 8 samples is most likely noise or an outlier, so it is dropped.

## Convert ENSEMBL to SYMBOL

Before we do any analysis, we need to do a little bit of a cleanup.

First we need to convert the ENS to SYMBOL.
We will use `annotables` package, that provides per gene information for different genome versions.
`airway` dataset uses Grch37 genome assembly.

```{r}
head(grch37)
```

We will use this table to build our own gene annotation table.

```{r}
gene_anno <- data.frame(ensgene = rownames(raw_counts)) %>%
  left_join(grch37, by= "ensgene") %>%
  distinct(symbol, .keep_all = TRUE) %>% # sometimes one ENSEMBL might have two different SYMBOLs or ENTREZ. It might happen where there are multiple isoforms of one gene etc. This makes sure that when an ENS is doubled, we only take the first observation.
  rename(gene_start = start, gene_end = end, gene_strand = strand) # start/end/strand are reserved names in rowData (GRanges), so DESeq2 would error

```

## Retain only protein coding genes

As you can see there are more than 60,000 genes in our dataset (before filtering).
Human genome has around 20,000 genes, so how is it possible?
Let's investigate `biotype` column

```{r}
table(gene_anno$biotype)
```

A lot of the transcripts are not even protein coding genes.
Lots of them are non-coding RNA, rRNA etc.
In this analysis we only care about protein coding genes.

First we will subset `gene_anno` to only protein coding genes and use that table to subset.

```{r}
gene_anno <- gene_anno %>%
  filter(biotype == 'protein_coding')
dim(gene_anno)
```

```{r}
raw_counts <- raw_counts[gene_anno$ensgene, ] # retain only protein coding genes
rownames(raw_counts) <- gene_anno$symbol # replace ENS with symbols count table
rownames(gene_anno) <- gene_anno$symbol
```

```{r}
dim(raw_counts)
```

```{r}
head(raw_counts)
```

We are left with \~13,000 genes

# DESeq2 workflow: Part I

We prepared the metadata and count matrix.
We are ready to create a deseq object (`dds`).
IT IS IMPORTANT THAT THE COUNTS ARE RAW!
If the counts are integers, they are most likely raw counts.

## Design

The **design** is a formula that tells DESeq2 which factors may explain differences in gene expression between your samples.
It is written as `~ variable1 + variable2`, where each variable is a column name from the metadata table - just like linear regression.

**Why is it important?** DESeq2 fits a model for every gene using the design.
It uses that model to separate the effect you care about (here, the drug) from other sources of variation (here, the cell line).
If the design is wrong or incomplete, the variation you did not account for stays in the noise.
That can hide real differences or create false ones.
The design cannot be fixed after the fact without re-running the analysis, so it is worth getting right before you create `dds`.

**How do we decide what goes there?**

1.  Start from your biological question. We want to know which genes change with dexamethasone, so `dex` goes in.
2.  Add the other variables that are known to affect expression and that you want to correct for, such as cell line, batch, sex, or age. Here the same four cell lines appear both treated and untreated, and different cell lines naturally express genes differently, so `cell` goes in. This is called paired design.
3.  Put the variable of interest **last**. DESeq2 tests the last variable in the formula by default.
4.  Only use columns that exist in the metadata, and only the ones that matter. Identifiers (sample name, run ID) are different for every sample and must not go in the design. Variables that are constant (like `albut` here) carry no information. Don't add variables just because you have them, since each one uses up samples that could have gone into the real comparison.

So our design is `~ cell + dex`.
In plain words: "compare treated and untreated samples, but control for individual effects of each cell line").

**Paired design:** the same cell line (or patient, animal, etc.) is measured in both conditions, so every treated sample has an untreated "partner".

**Why control for it?** Cell lines differ from each other far more than we expect treatment to change a typical gene.
If `cell` is not in the design, that cell-to-cell variation lands in the noise, so real drug effects are harder to detect.
With `cell` in the design, we account for these cell-line specific effects, focusing on the drug effect.

## Create a DESeq object

Let's relevel the factors in `dex` column to make sure that untrt is the reference.

```{r}
mtd$dex <- factor(mtd$dex, levels = c('untrt', 'trt'))
```

```{r, warning=FALSE}
dds <- DESeqDataSetFromMatrix(countData = raw_counts, 
                              colData = mtd, 
                              rowData = gene_anno,
                              design = ~ cell + dex)
```

```{r}
dds
```

dds object contains all this individual information and tables in one object.
Each element can be accessed

```{r}
dds@design
head(dds@colData) # metadata
head(counts(dds)) # counts
head(rowData(dds)) # gene_anno
```

## Running DESeq

`DESeq()` this function does the following steps:

1.  estimation of size factors - normalization
2.  estimation of dispersion
3.  Negative Binomial GLM fitting and Wald statistics - prepares for differential expression

These steps are basically normalization and preparation for differential expression testing (log2 fold change and p-value) and are almost always run with default parameters.

```{r}
dds <- DESeq(dds)
```

```{r}
dds
```

You can notice that after running this function, there are new elements added to the dds object: new assays, sizeFactor in colData etc.
You can also access normalized counts.
Notice how they changed from integers to decimals.

```{r}
head(counts(dds, normalize=TRUE))
# raw counts can still be obtained with
# counts(dds, normalize=FALSE)
```

# Principal component analysis

After RNA-seq you have a matrix of counts: 13,000 genes (rows) by 8 samples.
Each sample is a point in 13,000-dimensional space, which you can't look at.
But many genes move together.
For example, hundreds of inflammation genes rise and fall as a group.
So the data really vary along far fewer independent directions than 13,000.

PCA finds new axes, called principal components, that are weighted combinations of genes.
PC1 is the single direction along which the samples spread out the most.
PC2 is the next-most-variable direction, at right angles to PC1, and so on.
Plotting each sample's position on PC1 against PC2 gives a scatterplot.
Samples close together have similar overall expression profiles.
The axis labels report the percentage of total variance each PC captures, which tells you how much of the full picture the plot actually shows.

PCA is a quality-control and exploration step.
It is *not* part of the differential expression test itself.
You use it to ask a few questions:

- Do replicates of the same condition cluster together?

- Does PC1 separate your biological groups (treated vs. control), or does it separate something unwanted, like sequencing batch, sex, or the day samples were processed?
  If batch dominates, you'd add it to your design formula (`~ batch + condition`).

- Is there an outlier sample that looks like neither group?

PCA requires data transformation that is different that what we did with `DESeq()` function.
We need to apply variance-stabilizing transformation, which puts genes on a comparable scale.
To do so, we will use `vst()` function.

```{r}
vsd <- vst(dds, blind=FALSE)
```

```{r}
vsd
```

I will show two methods of running PCA: one is built in DESeq2 and is simpler, and another using `PCAtools` package, which I prefer because it gives me more control.

1.  DESeq default

```{r, warning=FALSE}
plotPCA(vsd, intgroup=c("cell", "dex"))
```

There is a message saying `using ntop=500 top features by variance`.
Most of the genes are not highly variable across samples (e.g. house keeping genes) and they are not very informative of sample composition and sample-to-sample differences.
That's why we use a fraction genes with the highest variance - they are the most informative ones, because they vary the most across samples.
`plotPCA` uses top 500 most highly variable genes by default.

2.  Using `PCAtools`

```{r, warning=FALSE}
pca_plot <- vsd %>%
  assay %>%
  pca(., vsd@colData, removeVar = 0.9) %>% # remove 90% of lower variance -> retains top 10% most variable genes
  biplot(colby = "cell",  shape = 'dex', legendPosition = 'right', lab = NULL )

pca_plot
```

Color represents cell line and shape represents treatment status.
Thanks to this we can easily identify the main sources of variation:

- PC1 explains 50% of variation and is driven by dex (treatment status) - triangles and dots are clearly separated on the X axis.
  This is great, because it means that the treatment made a real difference

- PC2 explains 20% variation and is driven by cell line - different colors separate along the Y axis, with the blue color separating the most from the rest.
  It is expected to see such spread between biological replicates, because each replicate is a separate cell line.
  There are differences in baseline expression between different cell lines, which is why we see such separation on PC2.
  This is also why we included cell line in the design - we don't want the differences between cell lines mask the effect of treatment, especially in a paired design.

We can color or shape the plot by every variable that exist in our metadata column.
It is a good practice to play with them, because it might help identify sources of technical variability that can be accounted for in the DESeq design.

> **Exercise 1**
>
> Using information available in the metadata table, check whether there are any other factors that can be correlated with PC1 or PC2.
> Write a short comment on your observations.

# DESeq2 workflow: Part II

## Differential expression test

Now that we normalized the data and did a QC check with PCA, we can perform Differential Gene Expression - the clue of the analysis.
In DGE we look for differences in gene expression across two different conditions.
For example: Gene A is 2x highly expressed in the treated group than untreated.
We will use `results()` function that return a table with log2 fold change difference between group 1 and group 2 and a p-value.
We need to specify `contrast` which takes a vector of characters: `c(column in metadata, group 1, group 2)`.
Here, we want to investigate the effect of dexamethasone treatment, so we will use dex column and compare trt vs. untrt.
Alternatively, we can use `name` parameter.
The names of coefs can be checked with `resultsNames`.
Sometimes, in complex designs, not all comparisons are generated.
In that case we can always stick to `contrast`

```{r}
resultsNames(dds)
```

```{r}
res <- results(dds, name = "dex_trt_vs_untrt", alpha = 0.05) # alpha is the p-value threshold to whether call a gene significant or not.

# alternatively (gives the same numbers, but lfcShrink() cannot be given this res, because it needs the coefficient name)
# res <- results(dds, contrast = c("dex", "trt", "untrt"), alpha = 0.05)
```

We can obtain a summary of our results

```{r}
summary(res)
```

Out of 13055 tested genes 2023 are expressed higher after the treatment than in control (upregulated) and 1810 are expressed lower after the treatment than in the control (downregulated)

```{r}
head(res)
```

`res` is the main result of the analysis.
It has one row per gene and answers the question: "did this gene change after dexamethasone treatment, and by how much?" The columns are:

- `baseMean` - average normalized expression of the gene across all samples. Tells you how highly the gene is expressed overall.
- `log2FoldChange` - the size and direction of the change (treated vs. untreated). A value of 1 means 2x higher in treated, -1 means 2x lower, 0 means no change.
- `lfcSE` - standard error of the log2 fold change, i.e. how uncertain that estimate is.
- `stat` - the Wald test statistic (`log2FoldChange / lfcSE`). The p-value is derived from it.
- `pvalue` - the probability of seeing a change at least this large if the gene was **not** truly affected by the treatment.
- `padj` - the p-value adjusted for multiple testing. **This is the value to use to call genes significant.**

## Adjusted p-value and multiple testing correction

A p-value of 0.05 means that there is a 5% chance of getting a result this extreme just by luck, even if nothing is going on.
That is fine for one test, but here we test \~13,000 genes at once.
If none of the genes were truly changed, we would still expect 0.05 × 13,000 = **\~650 genes** with p \< 0.05 purely by chance.
These false positives would be mixed in with the real hits, and we could not tell which are which.

```{r}
sum(res$pvalue < 0.05)
```

If we hadn't correct the p-value for multiple testing, we would have to expect that 13% of "significant" genes could be actually not significant - false positive

**Multiple testing correction** fixes this by making the threshold stricter the more genes we test.
DESeq2 uses the Benjamini-Hochberg method, which controls the **false discovery rate (FDR)**.
The adjusted p-value (`padj`) is the expected proportion of false positives among the genes called significant at that cutoff.
For example, if we keep all genes with `padj` \< 0.05, we accept that about 5% of them may be false positives.

```{r}
0.05 * sum(res$padj < 0.05)
```

A useful metric or test to check if our differential expression test was well calibrated and the results are reliable is to look the distribution of nominal (raw) p-values

- If the test is well calibrated, genes with no real effect (nulls) have p-values spread evenly between 0 and 1.
  Genes with a real effect pile up near 0.
  A healthy histogram is therefore a flat floor with a spike at the left edge.
  The height of the flat part gives a rough sense of how many genes are truly unchanged.expected, before you trust the padj values.

- Uniform shape with no spike: there is little or no signal, or the experiment is underpowered.

- Hill-shaped or skewed toward 1: the test is too conservative.
  This often points to a wrong variance or dispersion estimate, or to a problem in the pre-filtering.

- A spike near 0 plus an excess of p-values near 1: the model is probably misspecified, for example a missing batch or covariate in the design.

- Strange spikes at specific values: these usually come from low-count genes with discrete p-values.
  Filtering low-count genes (which the workflow already does) removes most of them.

Why it matters.
Multiple testing correction, including Benjamini-Hochberg, assumes the null p-values are uniform.
If they aren't, padj and the FDR it promises are unreliable, and the list of significant genes may contain many more false positives, or miss many more real ones, than stated.

```{r, fig.width=3, fig.height=3, warning=FALSE, message=FALSE}
p_hist <- ggplot(res, aes(x=pvalue)) +
  geom_histogram() +
  theme_bw()

p_hist
```

In our case the histogram looks perfect - our test is well-calibrated.

## LFC shrinking

`lfcShrink()` replaces the noisy maximum-likelihood log2 fold changes from `DESeq()` with empirical Bayes estimates pulled toward zero.
For genes with low counts or high dispersion, the raw estimates can be large by chance, and shrinkage reduces them in proportion to how little information the gene carries.
Genes with strong evidence barely move.

It is worth using whenever fold change magnitude matters: volcano and MA plots, ranking genes by LFC (including GSEA inputs), and applying an LFC cutoff.

```{r}
res <- lfcShrink(dds, coef = 'dex_trt_vs_untrt', res = res ,type='apeglm')
```

## Volcano Plot

Volcano plots are a common way to display DGE results.
Each point is a gene, with log2 fold change on the x-axis and -log10 adjusted p-value on the y-axis, so genes with large, significant changes sit in the upper corners.

DESeq2 doesn't have any built-in volcano plot function.
We can use `EnhancedVolcano` package that offers quick plotting functions.
I do recommend writing your own custom plotting functions.
First of all, you have more control of what is shown and the aesthetics, but it also help you understand the data structure better.

Let's test EnhancedVolcano first.

```{r, warning=FALSE, fig.width=5, fig.height=6}
EnhancedVolcano(res,
                lab = rownames(res),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Dex: trt vs. untrt', subtitle = NULL,
                pCutoff = 0.05,
                FCcutoff = 0,
                pointSize = 1.0)
```

**Let's create a custom volcano plot**

1.  Customize the `res` table.
    a.  add `diffexpressed` column with values: UP, DOWN, NS

```{r}
padj_cutoff <- 0.05
lfc_cutoff <- 0

deg_table <- res %>%
  data.frame() %>%
  mutate(geneid = rownames(.)) %>%
  filter(!is.na(padj)) %>%
  mutate(
    diffexpressed = case_when(padj > padj_cutoff ~ "NS",
                              log2FoldChange > lfc_cutoff & padj < padj_cutoff ~ "UP",
                              log2FoldChange < -lfc_cutoff & padj < padj_cutoff ~ "DOWN",
                              TRUE ~ "NS"))
```

```{r}
head(deg_table)
table(deg_table$diffexpressed)
```

2.  Built a volcano plot with ggplot2

```{r, warning=FALSE}
ggplot(deg_table, aes(x=log2FoldChange, y=-log10(padj), color = diffexpressed)) +
  geom_point(size = 1) +
  geom_vline(xintercept = lfc_cutoff, linetype= 'dashed', color = 'gray') +
  geom_hline(yintercept = padj_cutoff,linetype= 'dashed', color = 'gray') +
  scale_color_manual(values = c(DOWN = '#2E6DB4', NS = "gray40", UP="#D6403A"))+ 
  theme_bw() +
  theme(panel.grid = element_blank()) +
  coord_cartesian() +
  ggtitle('Dex: trt vs. untrt')
```

> **Exercise 2**
>
> Modify the volcano plot, by adding label names to the top 20 differentially expressed genes (DEGs), based on the adjusted p-value (hint: use ggrepel).

> **Exercise 3**
>
> Pick 5 DEGs and try to find basic information about their function.
> Given the context of this study, is it expected to see these genes being differentially expressed?

> **Exercise 4**
>
> You will run the DGE analysis many times and some of the steps might get repetitive.
> Write your own functions:
>
> Function 1 should take deseq result object, with arguments: `pvalue_cutoff`, `lfc_cutoff`, `ntop_deg_labels` ,and return a `data.frame` with additional column "diffexpressed" (UP, DOWN, NS)
>
> Function 2 should take the modified result table and return a ggplot volcano plot.
> Make it personalized for your own liking, but also flexible, e.g. optionally show top DEGs labels, how many genes to label etc.; display the number of UP, DOWN and NS genes.
>
> You can name the functions and arguments however you like, but make sure it is descriptive and other people can understand your code.

# Pathway enrichment analysis

An RNA-seq experiment gives you hundreds or thousands of differentially expressed genes.
Reading them one by one is not practical.
Gene Ontology (GO) helps by grouping genes into labeled sets, such as “inflammatory response” or “cell cycle”, based on what past studies reported about each gene.

Enrichment analysis asks whether your gene list contains more genes from a given set than you would expect by chance.
Say 5% of all expressed genes belong to “inflammatory response”.
If 20% of your upregulated genes do, that term is enriched.
The test is usually a hypergeometric (Fisher) test against a background of all genes you could have detected, followed by multiple-testing correction (FDR).

**Why we split up and down genes**\
Up and down genes usually reflect different biology.
In inflamed gut tissue, for example, upregulated genes tend to hit immune terms, and downregulated genes tend to hit epithelial transport and metabolism.
If you pool them, the larger group dominates and the smaller signal gets diluted.
Separate analyses are also easier to read, since each result list has a direction.
Splitting reduces the number of genes per test, which lowers power, so a weak signal can disappear (see Hong et al. 2014, *J R Soc Interface*, for a comparison of the two approaches).

**What a significant term means**\
A significant term means your up (or down) genes contain more members of that term than chance predicts.
That is all.
It is an association between a gene list and a label.

It does not mean the pathway is activated or repressed.
Several things can break that inference

- GO terms are unordered gene sets, with no direction.
  “Negative regulation of apoptosis” contains both genes that block cell death and genes that respond to it.

- Upregulated inhibitors can mean the pathway is switched on (negative feedback) or switched off.

- mRNA level is a proxy.
  Pathway activity often depends on protein abundance, phosphorylation, or localization, which RNA-seq does not measure.

- In bulk tissue, “T cell activation” going up can mean more T cells are present, not that each T cell is more active.
  Cell composition changes look like pathway changes.

- Well-studied genes carry more annotations, so popular terms get enriched more easily.

- Many GO terms overlap heavily (parent and child terms).
  Ten significant terms may be one signal counted ten times.

- The result depends on your cutoffs and background.
  Changing the padj or log2FC threshold, or using all genes instead of expressed genes as background, can change the top terms.

**How to use the results**\
Treat enriched terms as hypotheses and a way to generate a story, then test them.
Rank-based methods like GSEA avoid hard cutoffs.
Activity-inference tools (e.g. PROGENy, decoupleR) estimate pathway or TF activity from target gene behavior.
Orthogonal data such as protein levels, perturbation experiments, or single-cell data can confirm that the pathway really changed.\

We will perform overrepresantation test using `clusterProfiler` package, focusing on Gene Ontology: Biological Processes.
Make sure to install [org.Hs.eg.db](BiocManager::install(%22org.Hs.eg.db%22))

```{r}
# Define gene sets
# background genes = all tested genes
universe <- rownames(deg_table)
down_genes <- deg_table[deg_table$diffexpressed == 'DOWN', 'geneid']
```

```{r}
GO_down <- enrichGO(down_genes, 
                  OrgDb = "org.Hs.eg.db", # Human database
                  keyType = "SYMBOL", 
                  ont = "BP",
                  universe = universe, 
                  qvalueCutoff = 0.05,
                  readable = TRUE) 
# Some GO terms are reduntant. simplify should help remove redundancy but it might be computationally demanding. 
GO_down_clean <- simplify(GO_down)
```

```{r}
head(as.data.frame(GO_down_clean))
```

clusterProfiler offers lots of tools, including some plotting function.
Check their full [vignette](https://yulab-smu.top/biomedical-knowledge-mining-book/index.html).
It is common to show a couple of the most enriched terms.

```{r, fig.height=5, fig.width=6}
dotplot(GO_down, 
        x='FoldEnrichment', 
        showCategory = 15) +
  ggtitle("GO:BP, DOWN")
```

> **Exercise 5**
>
> Make the same plot for the 'cleaned' results.
> Are there any differences?
> How many significant terms does each table have?

```{r}

```

> **Exercise 6**
>
> Repeat the `enrichGO` and `simplify` for the upregulated genes.
> Show both groups (up and down simplified) side-by-side, using `patchwork` package.
> Which group has more specific GO terms, that can be associated with a more direct response to treatment (consider the cell type and what dex does to it)

# Summary

Congrats!
You made it through your first differential gene expression analysis.
As a last exercise:

1.  Prepare a figure that combines all the plots you have generated (`patchwork`).
2.  Write a short description what each panel represents.
3.  Write general observations.
4.  Write conclusions. Restate the hypotheses and the research question. Can you confirm/reject the hypotheses and the research question?

# Additional analyses

Kidding, we are not done yet ;)

Like I mentioned many times, there's usually more than one way to analyze genomics data.
Sometimes you may wonder, what my filtering step should look like, what should I include in the deseq design, should I apply a log2FC cutoff.
And often times the answer is: you need to try different approaches and see which one works best.

## Log2 fold change cutoff

The adjusted p-value tells you whether a gene’s change is distinguishable from zero.
It says nothing about size.
With enough replicates or high counts, a 5% change can be highly significant, and such changes are often too small to matter biologically.
A log2FC cutoff keeps only genes with an effect large enough to care about.
A common choice is \|log2FC\| \> 1 (2-fold) or and 0.5 is also used.
Neither value has a principled basis, so pick one based on your system and say so in the methods.

> **Exercise 7**
>
> 1.  Pick a LFC cutoff and apply it to `res` object. Remember that you have created a function to transform the deseq result table - utilize that.
> 2.  Run the enrichment analysis on the dataset with applied log2FC
> 3.  Make a figure that shows volcano plot and GO term plots for both no LFC cutoff and LFC cutoff
> 4.  Write down your observations. Did the LFC cutoff helped making the GO terms more specific?

## Alternative design

The design that we used so far was `~cell + dex`.
We said that including cell line information is important, because we are dealing with a paired design.
But happens if we don't include that information in our design?

> **Exercise 8**
>
> 1.  Create a new `dds` object (different variable name) where you run `~dex` model.
> 2.  Run and plot PCA
> 3.  Run `results` (no LFC cutoff)
> 4.  Plot a histogram of nominal p-values.
> 5.  Make a scatter plot:
>     a.  x: LFC of `~cell + dex` model; y: LFC of `~dex` model.
>     b.  Comute correlation between x and y. Show the fitted curve and report correlation coefficient and $R^2$
>     c.  Color the genes based on whether they are significant in both, significant in x, significant in y
> 6.  Make a figure that shows p-value histogram and PCA for both models side by side and the scatter plot:
>     a.  Are the two PCAs different or the same? If they look the same, why that might be? (hint: check in the DESeq2 vignette whether vst and pca include the design information. is there a way to include that?)
>     b.  How do the two p-value distributions compare? Any concerns?
>     c.  How many DEG did each model identified?
>     d.  How many overlapping and model-specific DEGs?
>     e.  Did including the cell line information improved the analysis?

# Next steps

Good job!
You now have a solid foundation on differential gene expression analysis.
Let's test your skills and understanding further.
This time you will pick your own dataset to work on.

Think of a couple of biological topics that you are interested in.
It can be a disease model of interest, particular organ, exposure to different environmental factors etc.
To make things easier, you can use AI (e.g. claude) to help you identify the study.
Describe your interests and ask it to search for studies with publicly available bulk RNA-seq, where raw count matrix and metadata/sample info are available, the design is relatively simple and the sample size is not gigantic.

Once you have a list of candidates, talk to me and we will pick the dataset together.
Your task will be to analyze the data: differential expression and functional enrichment analysis.
