---
title: "DESEq2 workflow"
output: html_notebook
---

# DESEq2 vignette

If you have any questions, check the [DESEq2 vignette](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html).

# About the dataset

This workflow uses the **airway** dataset (Bioconductor package `airway`), an RNA-seq experiment from [Himes et al. 2014, *PLoS ONE*](https://doi.org/10.1371/journal.pone.0099625) (GEO: [GSE52778](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE52778)).
The code used to generate the tables can be found [data/airway_generate_counts.R](https://github.com/wiktormiko/Intro-to-genomics/blob/main/vignettes/data/airway_generate_counts.R)

- **Biological question:** how do human airway smooth muscle (ASM) cells respond to dexamethasone, a glucocorticoid used to treat asthma and inflammation?
- **Design:** four ASM cell lines (`N61311`, `N052611`, `N080611`, `N061011`), each either left untreated (`untrt`) or treated with dexamethasone (`trt`).
- **Size:** 8 samples (4 cell lines x 2 conditions) and 63,677 genes.
- **Count matrix** (`airway_raw_counts.csv`): rows are genes (Ensembl IDs), columns are samples (SRA run IDs, e.g. `SRR1039508`). Values are raw, unnormalized read counts.
- **Metadata** (`airway_metadata.csv`): one row per sample.
  - `cell`: the cell line.
  - `dex`: dexamethasone treatment (`trt` / `untrt`). This is the main variable of interest.
  - `albut`: albuterol treatment. Every sample here is `untrt`, so it carries no information.
  - `SampleName`, `Run`, `Experiment`, `Sample`, `BioSample`: GEO / SRA identifiers.
  - `avgLength`: average read length.

Full description:

```{r, message=FALSE, warning=FALSE}
Sys.setenv(LANGUAGE = "en") # keep R messages in English regardless of system language
library(airway)
vignette("airway")
```

Because each cell line appears in both conditions, the design is **paired**.
In the DESeq2 model, `cell` should be included as a covariate alongside `dex`.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(annotables)
library(PCAtools)
```

# Import the data

```{r}
raw_counts <- read.csv("data/airway_raw_counts.csv", row.names = 1)
mtd <- read.csv("data/airway_metadata.csv", row.names = 1)
```

```{r}
head(raw_counts)
```

Count matrix is a table where rows correspond to gene names and columns - to samples.
The gene names are encoded in "ENSEMBL" system.
We will convert those names to SYMBOLs - regular gene names.

```{r}
head(mtd)
```

Metadata contains information about samples.
Each row is a sample and each column describes a metric.
For example, we can put information about treatment, batch, collection date, patient ID, etc.

It is important that the count table and metadata are matching.
Column names of the count matrix and row names of the metadata have to be the same!

We can ensure those two are matching by running this:

```{r}
all(rownames(mtd) == colnames(raw_counts))
```

if they're not in the right order we can use `match()` to reorder one table so it lines up with the other:

```{r}
raw_counts <- raw_counts[, match(rownames(mtd), colnames(raw_counts))]
```

# Pre–filtering

## Remove low count genes

It is not necessary to filter out low count genes, because DESEq pipeline does it automatically (`independent filtering`), but applying a small cutoff might make the object smaller and require less computational power and memory.
This cutoff is taken from the vignette.
It's also common to use `filterByExp` from edgeR package

`smallestGroupSize` - we have 4 treated and 4 controls so we will use 4.

```{r}
table(mtd$dex)
```

```{r}
ngenes_before <- nrow(raw_counts)
```

```{r}
smallestGroupSize <- 4
keep <- rowSums(raw_counts >= 10) >= smallestGroupSize
raw_counts <- raw_counts[keep,]
```

```{r}
ngenes_after <- nrow(raw_counts)
cat("#genes before filtering: ", ngenes_before, "\n #genes after filtering: ", ngenes_after)
```

1.  `raw_counts >= 10` gives a table of TRUE/FALSE with the same shape as the counts. Each cell asks whether this gene had at least 10 reads in this sample.
2.  `rowSums(...)` adds up the TRUE values for each gene, giving the number of samples where the gene reached 10 reads.
3.  `>= smallestGroupSize` asks whether that number is at least the size of your smallest experimental group.

A gene that is truly different between conditions should have decent counts in all the samples of at least one group.
If your smallest group has 4 samples, requiring at least 4 samples with 10 or more reads means a gene could pass by being expressed in just one condition, such as a gene that is switched on only after treatment.
A gene with high counts in only 1 of 8 samples is most likely noise or an outlier, so it is dropped.

## Convert ENSEMBL to SYMBOL

Before we do any analysis, we need to do a little bit of a cleanup.

First we need to convert the ENS to SYMBOL.
We will use `annotables` package, that provides per gene information for different genome versions.
`airway` dataset uses Grch37 genome assembly.

```{r}
head(grch37)
```

We will use this table to build our own gene annotation table.

```{r}
gene_anno <- data.frame(ensgene = rownames(raw_counts)) %>%
  left_join(grch37, by= "ensgene") %>%
  distinct(symbol, .keep_all = TRUE) %>% # sometimes one ENSEMBL might have two different SYMBOLs or ENTREZ. It might happen where there are multiple isoforms of one gene etc. This makes sure that when an ENS is doubled, we only take the first observation.
  rename(gene_start = start, gene_end = end, gene_strand = strand) # start/end/strand are reserved names in rowData (GRanges), so DESeq2 would error

```

## Retain only protein coding genes

As you can see there are more than 60,000 genes in our dataset (before filtering).
Human genome has around 20,000 genes, so how is it possible?
Let's investigate `biotype` column

```{r}
table(gene_anno$biotype)
```

A lot of the transcripts are not even protein coding genes.
Lots of them are non-coding RNA, rRNA etc.
In this analysis we only care about protein coding genes.

First we will subset `gene_anno` to only protein coding genes and use that table to subset.

```{r}
gene_anno <- gene_anno %>%
  filter(biotype == 'protein_coding')
dim(gene_anno)
```

```{r}
raw_counts <- raw_counts[gene_anno$ensgene, ] # retain only protein coding genes
rownames(raw_counts) <- gene_anno$symbol # replace ENS with symbols count table
rownames(gene_anno) <- gene_anno$symbol
```

```{r}
dim(raw_counts)
```

```{r}
head(raw_counts)
```

We are left with \~13,000 genes

# DESeq2 workflow: Part I

We prepared the metadata and count matrix.
We are ready to create a deseq object (`dds`).
IT IS IMPORTANT THAT THE COUNTS ARE RAW!
If the counts are integers, they are most likely raw counts.

## Design

The **design** is a formula that tells DESeq2 which factors may explain differences in gene expression between your samples.
It is written as `~ variable1 + variable2`, where each variable is a column name from the metadata table - just like linear regression.

**Why is it important?** DESeq2 fits a model for every gene using the design.
It uses that model to separate the effect you care about (here, the drug) from other sources of variation (here, the cell line).
If the design is wrong or incomplete, the variation you did not account for stays in the noise.
That can hide real differences or create false ones.
The design cannot be fixed after the fact without re-running the analysis, so it is worth getting right before you create `dds`.

**How do we decide what goes there?**

1.  Start from your biological question. We want to know which genes change with dexamethasone, so `dex` goes in.
2.  Add the other variables that are known to affect expression and that you want to correct for, such as cell line, batch, sex, or age. Here the same four cell lines appear both treated and untreated, and different cell lines naturally express genes differently, so `cell` goes in. This is called paired design.
3.  Put the variable of interest **last**. DESeq2 tests the last variable in the formula by default.
4.  Only use columns that exist in the metadata, and only the ones that matter. Identifiers (sample name, run ID) are different for every sample and must not go in the design. Variables that are constant (like `albut` here) carry no information. Don't add variables just because you have them, since each one uses up samples that could have gone into the real comparison.

So our design is `~ cell + dex`.
In plain words: "compare treated and untreated samples, but control for individual effects of each cell line").

**Paired design:** the same cell line (or patient, animal, etc.) is measured in both conditions, so every treated sample has an untreated "partner".

**Why control for it?** Cell lines differ from each other far more than we expect treatment to change a typical gene.
If `cell` is not in the design, that cell-to-cell variation lands in the noise, so real drug effects are harder to detect.
With `cell` in the design, we account for these cell-line specific effects, focusing on the drug effect.

## Create a DESeq object

```{r, warning=FALSE}
dds <- DESeqDataSetFromMatrix(countData = raw_counts, 
                              colData = mtd, 
                              rowData = gene_anno,
                              design = ~ cell + dex)
```

```{r}
dds
```

dds object contains all this individual information and tables in one object.
Each element can be accessed

```{r}
dds@design
head(dds@colData) # metadata
head(counts(dds)) # counts
head(rowData(dds)) # gene_anno
```

## Running DESeq

`DESeq()` this function does the following steps:

1.  estimation of size factors - normalization
2.  estimation of dispersion
3.  Negative Binomial GLM fitting and Wald statistics - prepares for differential expression

These steps are basically normalization and preparation for differential expression testing (log2 fold change and p-value) and are almost always run with default parameters.

```{r}
dds <- DESeq(dds)
```

```{r}
dds
```

You can notice that after running this function, there are new elements added to the dds object: new assays, sizeFactor in colData etc.
You can also access normalized counts.
Notice how they changed from integers to decimals.

```{r}
head(counts(dds, normalize=TRUE))
# raw counts can still be obtained with
# counts(dds, normalize=FALSE)
```

# Principal component analysis

After RNA-seq you have a matrix of counts: 13,000 genes (rows) by 8 samples.
Each sample is a point in 13,000-dimensional space, which you can't look at.
But many genes move together.
For example, hundreds of inflammation genes rise and fall as a group.
So the data really vary along far fewer independent directions than 13,000.

PCA finds new axes, called principal components, that are weighted combinations of genes.
PC1 is the single direction along which the samples spread out the most.
PC2 is the next-most-variable direction, at right angles to PC1, and so on.
Plotting each sample's position on PC1 against PC2 gives a scatterplot.
Samples close together have similar overall expression profiles.
The axis labels report the percentage of total variance each PC captures, which tells you how much of the full picture the plot actually shows.

PCA is a quality-control and exploration step.
It is *not* part of the differential expression test itself.
You use it to ask a few questions:

- Do replicates of the same condition cluster together?

- Does PC1 separate your biological groups (treated vs. control), or does it separate something unwanted, like sequencing batch, sex, or the day samples were processed?
  If batch dominates, you'd add it to your design formula (`~ batch + condition`).

- Is there an outlier sample that looks like neither group?

PCA requires data transformation that is different that what we did with `DESeq()` function.
We need to apply variance-stabilizing transformation, which puts genes on a comparable scale.
To do so, we will use `vst()` function.

```{r}
vsd <- vst(dds, blind=FALSE)
```

```{r}
vsd
```

I will show two methods of running PCA: one is built in DESeq2 and is simpler, and another using `PCAtools` package, which I prefer because it gives me more control.

1.  DESeq default

```{r, warning=FALSE}
plotPCA(vsd, intgroup=c("cell", "dex"))
```

There is a message saying `using ntop=500 top features by variance`.
Most of the genes are not highly variable across samples (e.g. house keeping genes) and they are not very informative of sample composition and sample-to-sample differences.
That's why we use a fraction genes with the highest variance - they are the most informative ones, because they vary the most across samples.
`plotPCA` uses top 500 most highly variable genes by default.

2.  Using `PCAtools`

```{r, warning=FALSE}
vsd %>%
  assay %>%
  pca(., vsd@colData, removeVar = 0.9) %>% # remove 90% of lower variance -> retains top 10% most variable genes
  biplot(colby = "cell",  shape = 'dex', legendPosition = 'right', lab = NULL )
```

Color represents cell line and shape represents treatment status.
Thanks to this we can easily identify the main sources of variation:

- PC1 explains 50% of variation and is driven by dex (treatment status) - triangles and dots are clearly separated on the X axis.
  This is great, because it means that the treatment made a real difference

- PC2 explains 20% variation and is driven by cell line - different colors separate along the Y axis, with the blue color separating the most from the rest.
  It is expected to see such spread between biological replicates, because each replicate is a separate cell line.
  There are differences in baseline expression between different cell lines, which is why we see such separation on PC2.
  This is also why we included cell line in the design - we don't want the differences between cell lines mask the effect of treatment, especially in a paired design.

We can color or shape the plot by every variable that exist in our metadata column.
It is a good practice to play with them, because it might help identify sources of technical variability that can be accounted for in the DESeq design.

# DESeq2 workflow: Part II

## Differential expression test

Now that we normalized the data and did a QC check with PCA, we can perform Differential Gene Expression - the clue of the analysis.
In DGE we look for differences in gene expression across two different conditions.
For example: Gene A is 2x highly expressed in the treated group than untreated.
We will use `results()` function that return a table with log2 fold change difference between group 1 and group 2 and a p-value.
We need to specify `contrast` which takes a vector of characters: `c(column in metadata, group 1, group 2)`.
Here, we want to investigate the effect of dexamethasone treatment, so we will use dex column and compare trt vs. untrt

```{r}
deg_table <- results(dds, contrast = c("dex", "trt", "untrt"), alpha = 0.05) # alpha is the p-value threshold to whether call a gene significant or not. 
```

We can obtain a summary of our results

```{r}
summary(deg_table)
```

Out of 13055 tested genes 2023 are expressed higher after the treatment than in control (upregulated) and 1810 are expressed lower after the treatment than in the control (downregulated)

```{r}
head(deg_table)
```

`deg_table` is the main result of the analysis.
It has one row per gene and answers the question: "did this gene change after dexamethasone treatment, and by how much?" The columns are:

- `baseMean` - average normalized expression of the gene across all samples. Tells you how highly the gene is expressed overall.
- `log2FoldChange` - the size and direction of the change (treated vs. untreated). A value of 1 means 2x higher in treated, -1 means 2x lower, 0 means no change.
- `lfcSE` - standard error of the log2 fold change, i.e. how uncertain that estimate is.
- `stat` - the Wald test statistic (`log2FoldChange / lfcSE`). The p-value is derived from it.
- `pvalue` - the probability of seeing a change at least this large if the gene was **not** truly affected by the treatment.
- `padj` - the p-value adjusted for multiple testing. **This is the value to use to call genes significant.**

## Adjusted p-value and multiple testing correction

A p-value of 0.05 means that there is a 5% chance of getting a result this extreme just by luck, even if nothing is going on.
That is fine for one test, but here we test \~13,000 genes at once.
If none of the genes were truly changed, we would still expect 0.05 × 13,000 = **\~650 genes** with p \< 0.05 purely by chance.
These false positives would be mixed in with the real hits, and we could not tell which are which.

```{r}
sum(deg_table$pvalue < 0.05)
```

If we hadn't correct the p-value for multiple testing, we would have to expect that 13% of "significant" genes could be actually not significant - false positive

**Multiple testing correction** fixes this by making the threshold stricter the more genes we test.
DESeq2 uses the Benjamini-Hochberg method, which controls the **false discovery rate (FDR)**.
The adjusted p-value (`padj`) is the expected proportion of false positives among the genes called significant at that cutoff.
For example, if we keep all genes with `padj` \< 0.05, we accept that about 5% of them may be false positives.

```{r}
0.05 * sum(deg_table$padj < 0.05)
```

A useful metric or test to check if our differential expression test was well calibrated and the results are reliable is to look the distribution of nominal (raw) p-values

- If the test is well calibrated, genes with no real effect (nulls) have p-values spread evenly between 0 and 1.
  Genes with a real effect pile up near 0.
  A healthy histogram is therefore a flat floor with a spike at the left edge.
  The height of the flat part gives a rough sense of how many genes are truly unchanged.expected, before you trust the padj values.

- Uniform shape with no spike: there is little or no signal, or the experiment is underpowered.

- Hill-shaped or skewed toward 1: the test is too conservative.
  This often points to a wrong variance or dispersion estimate, or to a problem in the pre-filtering.

- A spike near 0 plus an excess of p-values near 1: the model is probably misspecified, for example a missing batch or covariate in the design.

- Strange spikes at specific values: these usually come from low-count genes with discrete p-values.
  Filtering low-count genes (which the workflow already does) removes most of them.

Why it matters.
Multiple testing correction, including Benjamini-Hochberg, assumes the null p-values are uniform.
If they aren't, padj and the FDR it promises are unreliable, and the list of significant genes may contain many more false positives, or miss many more real ones, than stated.

```{r}
hist(deg_table$pvalue)
```

In our case the histogram looks perfect - our test is well-calibrated.

## Volcano Plot

enhanced volcano and custom plot

```{r}
print(dds)
```
